---
title: "Clean Sentinel data"
output: html_document
date: "2023-01-26"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(rbeni)
source("R/init_dates_dataframe.R")
```

## Read file

```{r cars}
df <- read_csv("data/pixel_data.csv") |> 
  
  # clean date
  mutate(timestamp = lubridate::dmy(timestamp)) |> 
  
  # arrange by date
  arrange(timestamp) |> 
  
  rename(date = timestamp) |> 
  
  # get decimal year
  mutate(year_dec = lubridate::decimal_date(date))
```

Fraction of cloud contaminated
```{r}
df |> 
  ggplot(aes(cp, ..density..)) +
  geom_histogram()
```

## Clean data

Retaining only data with CP < 0.01.

```{r}
df_clean <- df |> 
  # clean based on cloud probability into classes
  mutate(qc = ifelse(cp < 0.01, "good", "bad")) |> 
  mutate(ndvi_clean = ifelse(qc == "good", ndvi, NA)) |> 
  drop_na(ndvi_clean)
```

Drop outliers per week-of-year. Outliers are determined here by R's `boxplot.stat()` function.

```{r}
remove_outliers <- function(df, varnam){
  outliers <- boxplot.stats(df[[varnam]])$out
  df[[varnam]] <- ifelse(df[[varnam]] %in% outliers, NA, df[[varnam]])
  return(df)
}

df_clean2 <- df_clean |> 
  mutate(week = lubridate::week(date)) |> 
  group_by(week) |> 
  nest() |> 
  mutate(data = purrr::map(data, ~remove_outliers(., "ndvi_clean"))) |> 
  unnest(data) |> 
  ungroup() |> 
  drop_na()
```

Create a data frame spanning all dates (daily, not just dates where Sentinel observations are available). The function `init_dates_dataframe()` is in subdirectory `./R/`.
```{r}
ddf <- init_dates_dataframe(yrstart = min(year(df$date)), yrend =  max(year(df$date))) |> 
  left_join(df |> 
              select(-year_dec), 
            by = "date")
```

## Smooth and interpolate time series

### Savitzky-Golay filter

```{r}
df_clean2 <- df_clean2 |> 
  mutate(ndvi_sg = signal::sgolayfilt(ndvi_clean, p = 7, n = 15, m = 0))

# linearly interpolate to daily output based on filtered
ddf$ndvi_sg <- stats::approx(df_clean2$year_dec, 
                                df_clean2$ndvi_sg, 
                                xout = ddf$year_dec)$y
```

### Spline

```{r}
mod_spline <- with(df_clean2, stats::smooth.spline(year_dec, ndvi_clean, spar = 0.01))
ddf$ndvi_spline = predict(mod_spline, ddf$year_dec)$y
```

### LOESS

```{r}
# determine periodicity of data (here 1 for one day)
period <- ddf |>
  mutate(prevdate = lag(date)) |>
  mutate(period = as.integer(difftime(date, prevdate))) |>
  pull(period) |>
  min(na.rm = TRUE)

# take a three-weeks window for locally weighted regression (loess)
# good explanation: 
# https://rafalab.github.io/dsbook/smoothing.html#local-weighted-regression-loess
ndays_tot <- lubridate::time_length(diff(range(ddf$date)), unit = "day")
span <- 100*period/ndays_tot

mod_loess <- stats::loess( ndvi_clean ~ year_dec, data = df_clean2, span = span)

ddf$ndvi_loess <- stats::predict( mod_loess, newdata = ddf )
```

### Linear interpolation

```{r}
ddf$ndvi_intpl <- stats::approx(df_clean2$year_dec, 
                                df_clean2$ndvi_clean, 
                                xout = ddf$year_dec)$y
```

## Plot full time series

```{r}
ggplot() +
  geom_point(aes(date, ndvi), color = "grey", data = df) +
  geom_point(aes(date, ndvi_clean), color = "green", data = df_clean) +
  geom_point(aes(date, ndvi_clean), color = "black", data = df_clean2) +
  geom_line(aes(date, ndvi_sg), color = "royalblue", data = ddf) +
  geom_line(aes(date, ndvi_loess), color = "orchid", data = ddf) +
  geom_line(aes(date, ndvi_intpl), color = "tomato", data = ddf) +
  ylim(-0.2, 1)
```

- Grey dots: removed based on cloud probability
- Green dots: removed based on outliers within week-of-year.
- Black dots: remaining good-quality data
- Blue line: Savitzky-Golay
- Red line: linear interpolation
- purple line: LOESS smoothing


## Years

Compare smoothed NDVI across years using, separate for each smoothing approach.

### Linearly interpoloated

```{r}
ddf |> 
  mutate(doy = lubridate::yday(date),
         year = as.factor(lubridate::year(date))) |> 
  ggplot(aes(x = doy, y = ndvi_intpl, group = year, color = year)) +
  geom_line() +
  ylim(-0.5, 1)
```

### LOESS

```{r}
ddf |> 
  mutate(doy = lubridate::yday(date),
         year = as.factor(lubridate::year(date))) |> 
  ggplot(aes(x = doy, y = ndvi_loess, group = year, color = year)) +
  geom_line() +
  ylim(-0.5, 1)
```

### Savitzky-Golay

Not very convincing!
```{r}
ddf |> 
  mutate(doy = lubridate::yday(date),
         year = as.factor(lubridate::year(date))) |> 
  ggplot(aes(x = doy, y = ndvi_sg, group = year, color = year)) +
  geom_line() +
  ylim(-0.5, 1)
```


